---
title: "A practical course in programming for biologists - Lecture 2"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

***

Lecture 2
Date: Friday, March 17, 2020

*** 

# Commands (or functions)
## Key Programming Concept

***

## Commands in R

>print("hello")

* **print** is the command
* **()** tells R that **print** is a command
* **"hello"** is the **argument** to the command
* These are also called **"functions”**

```{r}
print("Hello")
```

```{r}
cat("hello", "world")
```

***

# Commands in R

* **Arguments** can be passed flexibly  
Here's three different ways to find help:

```{r eval=FALSE, include=TRUE}
help(seq)
?seq
??seq
```

* Either follow the prescribed order (ok to omit ones at the end)

```{r}
seq(1,3,0.2)
```

* Specify arguments with "="

```{r}
seq(1,4,length.out=3)
```

* The order is not important if you specify with =
* `seq(to=5, from=10)` is the same as `seq(from=10, to=5)`
```{r}
seq(to=5, from=10)
seq(from=10, to=5)
```


***

# Variables
## Key Programming Concept

***

#Variables

* Variables are named places to store values

```{r}
x = 5
x
```

* Variable names
    * Syntax is key (as in most programming languages)
    * Letters, numbers, and the dot and underscore characters, and starts with either a letter or a dot not followed by a number.
    * Reserved words (e.g. TRUE, FALSE, NA, Inf, NaN) are not syntactic names
    * Some constants are already defined variables e.g. pi
* We **declare** variables in R by creating a new variable name
* You can use any name you want, as long as you follow the syntax rules

> Which of these are valid R variable names?  
name= "Tom"  
myname# = "Alice"  
_str="test"  
DNA="ATTTCA"  
sequence1="ATTTCA"  
.seq = "ATTTCA"  
1seq="ATTTCA"  
year@2008 = 2008  
gen2str = 1

# Variable assignment and types
* We assign a value to a variable using the assignment operator

```{r}
number = 5
number

number <- 5
number

string = "five"
string

number = number +1
number

try({
  string = string + 1
},
silent=T)
string
```

* All programming languages distinguish different variable types.
    * Numbers (integers, various non-integers)
    * Strings (=text, zero or more text characters)
    * Various others (e.g. vectors, matrices – later)

# Variable types

* Strictly speaking
    * Arithmetic operations are only defined for numbers
    * String operations are only defined for strings
* In practice, R is lenient and converts variables if possible

# Special characters

* Special characters have special meaning in the language
    * `~!@#$%^&*(){}_+:"<>?,./;'[]-=`
* " **starts** and **ends** a string
* \\ **escapes** a special character e.g. \\"
* Within a string:
    * \\n = newline
    * \\t = tab
* \# - starts a **comment** line and is ignored by R
    * `#my comment`
    * Best practice: Use these a lot to document your code

# Printing to the console/screen

```{r}
x
print(x) #equivalent to above
cat(x)   #similar to print, but less autoconversion

#Mixing variables within strings:
paste("Number", number, "String", string)

#Here is another way to do the same thing
sprintf("Number %s String %s", number, string)

#How do you print a quote?
string = "gene \"ABC\""

#Note:
print(string) #outputs `"gene \"ABC\""

cat(string)  #outputs gene "ABC"

#Here is a way to use print to get a similar result to cat in this case
print(string, quote=FALSE)

#You can use the \t string to denote a tab
cat("Separated\tby\ttabs")
```


# Exercise: Using variables (10 minutes)

* Use the R text editor to write a program that does the following:
    * Create a variable number1 and set it to a value
    * Do the same for number2
    * Compute the sum, product and ratio of these numbers
    * Print these to the screen
* Save your work as sum.r
* Run it in the R console

# Exercise solution (5 minutes)


***

# End of lecture 2

Reminder:
Complete Assignment 1 on Quercus
Due Monday, March 18, 2019 at 9am

Read The Book of R, section 2.3 (vectors), chapter 3 (matrices), 4.1 (logicals), 4.2 (characters/strings)

***
