---
title: "Module 2"
---


The Book of R: section 2.3 (vectors), chapter 3 (matrices), 4.1 (logicals), 4.2 (characters/strings)

## Module 1 review

* Introduction to the R statistical programming environment
* Commands/functions e.g. c(1, 2, 3)
* Variables
    * Name syntax (special characters), declaring, assigning
* Variable types
    * numeric, string, logical

*** 

# Sample code: numbers

```{r}
#Lots of things you can do with numeric variables
x = 2
x

y = 2*x-3
y

paste("x=",x," 2*x-3=",y)

x = x+1
x

x = x-1
x

x = x*1.5
x

y = x**2
y

x = y**(1/2)
x

z = exp(1)
z

x = z/2
x

paste("exp(1)/2 =", x)
```

# Operators

* Operators operate on variables
* = is the assignment operator
* Arithmetic operators
    * \+		addition
    * \- 		subtraction
    * \*		multiplication
    * /		division
    * ^ or **	exponentiation
* Comparison operators
    * <		less than
    * <=	less than or equal to
* And many more. See e.g. https://stat.ethz.ch/R-manual/R-devel/library/base/html/Arithmetic.html

# Sample code: strings

```{r}
# this is a string...
DNA = "ACGGGAGGACGGGAAAATTACTACGGCATTAGC"

#single vs. double quotes
s = "string123"
#s2 = "string"123"  #what's happening here?  Fix the code. (uncomment the line to see the error)
s3 = 'string123'
s4 = 'string"123'  #what's happening here?  Fix the code. (uncomment the line to see the error)

# concatenate (join) strings
s = paste(s, s4)
s = cat(s, s) # what is s?

#Note: cat and paste have different return values. You can see what a command returns in the help page.
cat(s,s)  #same thing as above cat command, but we don't need the return value, so we don't catch it in a variable. This would be the normal way to use the cat command.

# nchar() finds string length
n = nchar(DNA)

# substrings give you part of a string. Start counting at 1
alphabet = "abcdefghijklmnopqrstuvwxyz"
substr(alphabet,3,5)

#Substitution demo
sub("A", "T", DNA)
gsub("A", "T", DNA)

# what does the g mean?
sub("GA", "CT", DNA) # what happens if I include multiple characters in the pattern?

# If I want to save the output
DNA2=gsub("A", "T",DNA)

#Translation demo
chartr("A", "T", DNA)
chartr("ACGT", "AAAA", DNA) # how is this different from substitution?
```

# Example code: reverse_complement.r

```{r}
# Using string operations to answer a biological question
# What is the reverse complement of a DNA sequence?

DNA = readline("Please enter a DNA sequence. I will return the reverse complement.\n\n")

# Print the DNA onto the screen
cat("\nHere is the starting DNA sequence:\n")
cat(DNA)
cat("\n\n")

# Calculate the reverse complement
#  Warning: this attempt will fail!
#
# First, copy the DNA into new variable revcom
# (short for REVerse COMplement)
# It doesn't matter if we first reverse the string and then
# do the complementation; or if we first do the complementation
# and then reverse the string.  We'll get the same result each time.
# So when we make the copy we'll do the reverse in the same statement.
#

revcom = paste(rev(strsplit(DNA, NULL)[[1]]), collapse="")

#
# Next substitute all bases by their complements,
# A->T, T->A, G->C, C->G
#

revcom = gsub("A","T", revcom)
revcom = gsub("T","A", revcom)
revcom = gsub("G","C", revcom)
revcom = gsub("C","G", revcom)

# Print the reverse complement DNA onto the screen
cat ("Here is the reverse complement DNA:\n")
cat(revcom)

#
# Oh-oh, that didn't work right!
# Our reverse complement should have all the bases in it, since the
# original DNA had all the bases-but ours only has A and G!
#
# Do you see why?


cat("\n\nThat was a bad algorithm, and the reverse complement was wrong!\n")
readline("Press enter to try again\n\n")

# Make a new copy of the DNA (see why we saved the original?)
revcom = paste(rev(strsplit(DNA, NULL)[[1]]), collapse="")

revcom = chartr("ACGTacgt", "TGCAtgca", revcom)

#Print the reverse complement DNA onto the screen
cat ("Here is the reverse complement DNA:\n")
cat(revcom)
cat("\n\n")
cat("This time it worked!\n")
```

# Exercise: Reading in data from the console (10 minutes)

>Write a program that does the following:

* Prompt the user to input a number (number 1)
    * Hint: use the readline function
    * n = readline("What is the first number?\n")
    * Also, as.numeric(n) to convert to a number
* Prompt the user to input another number (number 2)
* Output the sum, product and ratio of these numbers

> Remember to use a code chunk in this R notebook to run your code

### Exercise solution (5 minutes)

```{r}
#Add your exercise code here. You can create your own code chunk by typing the text you see at the start "```{r}" and end "```" of the chunk, or by using the 'insert' button at the top of this editor tab

n1 <- readline("What is the first number?\n")
is(n1) #can we do math on this?
n1 <- as.numeric(n1) #coerce to numeric type

n2 <- as.numeric(readline("What is the second number?\n"))
# nesting functions works too.

print(n1 + n2)
print(n1 * n2)
print(n1 / n2)

```


# More variable types

## Vectors and Matrices

# Vectors: creating

* One of the variable types in R
* A vector is a sequence of data of the same type (e.g. numbers, strings)
* Create a vector with the `c()` function
```{r}
vector = c(1,2,3,4)
vector

x = c(1.2,2,54,6)
x

y = c(10,23.4,2,3)
y
```

# Vectors: math

```{r}
# We can multiply two vectors element by element
print(x*y)

# Can also do a dot product of two vectors
x %*% y

#Note: In the console, you don't need to use print()
```

# Vectors: types

* We can also create text or LOGICAL vectors

```{r}
textvec=c("Patient1","Patient2","Patient3")
textvec
```

```{r}
#Not compatible with arithmetic
#print(textvec * 2)  #gives an error
```


```{r}

logicvec=c(TRUE,FALSE,TRUE,FALSE)
logicvec

#Can also write:
c(T,F,T,F)
```

# Vectors: indexing

* Indexing is for accessing subsets of vectors
* For indexing, we always use the square brackets [ ]

```{r}

textvec[1]
textvec[3]

#We can also simultaneously access multiple elements
textvec[1:2]
textvec[c(1,1,3,3)]

#The length command
length(textvec)

```

# Vectors: names

```{r}
#R allows named elements
namedvec=c(firstcol=3, B=245, C=1212, lastelement=29.3)
namedvec

#We can then access the elements using the names
namedvec["firstcol"] * 10

#If we access an element with the wrong name we get NA
namedvec["asdf"]

#names() returns the vector of names
names(namedvec)
```

# Exercise: Vectors and indexing (10 minutes)

>Create a vector “radii” of length three containing the numbers 5, 10 and 20. Use “*” and the built-in constant “pi” to compute the areas of the three circles. Then subtract 2.1 and recompute the areas

>Use the “seq” and “sum” functions for this exercise - remember, you can always look up help using help(function): Generate a vector “counter” with a sequence of numbers from 1 to 100. Compute the sum of these numbers

>Generate a vector “randomdata” of length 100 with random numbers using the command “runif”

>Bonus: Make a vector sorter that will sort the random numbers using the “order” function and return the sorted random data in “sortedrandom”

### Exercise solution (5 minutes)

```{r,fig.width=8,fig.height=3,fig.show="hold"}
# step 1
radii <- c(5,10,20)
pi * radii * radii
pi * radii^2

# step 2
counter <- seq(1,100)
sum(counter)
sum(seq(1,100))

# step 3
randomstuff <- runif(100,min=1,max=100)
par(mfrow=c(1,2),mar=c(3,3,2,1),mgp=2:0)
hist(randomstuff) #runif randomly samples from the uniform distribution
hist(rnorm(100)) #rnorm randomly samples from the normal distribution


# step THE BONUS
sort(randomstuff,decreasing=F)
randomstuff[order(randomstuff,decreasing=F)]

identical(randomstuff[order(randomstuff,decreasing=F)],
          sort(randomstuff,decreasing=F))

```

# Matrices

>A matrix is like a 2D vector (i.e. a square or rectangle). Like a spreadsheet table with rows, columns

```{r}

#Create with the matrix command
firstmat = matrix(c(1,2.3,3,4,5.3,6.4),nrow=2,ncol=3,dimnames = list(c("row1","row2"),c("col1","col2","col3")))
firstmat

#Values are given in the first argument, the dimensions in the 2nd/3rd argument, dimension names in the last argument

#We can extract whole rows and columns like this
firstmat[,1]  #first column. firstmat[,2] – second column
firstmat[1,]  #first row

#We can also access rows and columns by name
firstmat[,"col2"]

#We can access single elements or subsets of the matrix like this
firstmat[1,2]
firstmat[1:2,2:3]

```

# Exercise: Matrices (10 minutes)

>Create a matrix with 3 columns and 10 rows

>Fill it with random numbers using the runif() command

>Calculate the average value in the matrix using the mean() command

### Exercise solution (5 minutes)

```{r}
ourrandommat <- matrix(runif(30),nrow=10,ncol=3)
mean(ourrandommat)
rowMeans(ourrandommat)
colMeans(ourrandommat)
```


***
