---
title: "Module 3"
---

The Book of R: chapter 3.4 (arrays), 5.1 (lists), 5.2 (data frames)

***

## Module 2 review

* Variable types: vector, matrix
* Vector
    * x = c(1.2,2,54,6)
    * textvec[1:2]
* Matrix
    * firstmat[1,2]
    * firstmat[1:2,2:3]
    
***

# Exercise: String commands (10 minutes)

> We used this code to reverse a string

>`paste(rev(strsplit(DNA, NULL)[[1]]), collapse="")`

>Figure out how this works by trying each of the nested functions, starting from `strsplit(DNA, NULL)`

>Note: `[[1]]` extracts elements within a list (which is a type of vector that can contain other lists)

### Exercise solution (5 minutes)

```{r}
DNA <- "ACGGGAGGACGGGAAAATTACTACGGCATTAGC"

str(strsplit(DNA,"")) # a list
str(strsplit(DNA,"")[1]) # still a list
str(strsplit(DNA,"")[[1]]) # a character vector (the first element of the list)

strsplit(DNA,"")[[1]]
rev(strsplit(DNA,"")[[1]]) #backwards!

paste(rev(strsplit(DNA,"")[[1]]),collapse="") # "collapsed" into a string

```


# Variables (continued)

## Lists, arrays, data frames

# Lists: creating

* Lists are vectors that contain a mix of variable types
* Create a list using the list() function

```{r}
numericVector = c(1, 3, 5)
numericVector
#Note: numericVector is just a variable name - we call it numericVector because we define it here to contain numbers, but it could be called any variable name that is allowed.

stringVector = c("abc", "def", "ghi")
stringVector

logicVector = c(T, F, T, F)
logicVector

listVariable = list(numericVector, stringVector, logicVector, 77)
listVariable
```

# Lists: accessing

```{r}
listVariable
listVariable[1]  #access first list element
listVariable[[1]] #access content of first element
listVariable[[1]][2] 

listVariable[[2]][1] = "ACGT"
listVariable[[2]]
listVariable[c(2, 4)]
```

# Lists: named elements

```{r}
listVariable = list(numbers=numericVector, strings=stringVector, logic=logicVector, myNum=77)
listVariable

listVariable["numbers"]
listVariable["myNum"]

#These are equivalent for named elements
listVariable[[1]]
listVariable[["numbers"]]
listVariable$numbers  #easiest access method

```

# Exercise: Lists (10 minutes)

* Create a list containing a vector of odd numbers from 1 to 100
    * Hint: you can use the seq() command
* Calculate the sum of this list
    * Hint: use the sum() command
* Repeat this, but name the vector and access its values from the list using its name

### Exercise solution (5 minutes)

```{r}
ourlist <- list(odds=seq(1,101,2),
                "evens"=seq(2,100,2),
                "some dumb words"=c("my","name","is"))
str(ourlist)
str(ourlist[1])
str(ourlist[[1]])

sum(ourlist[[1]])
sum(ourlist$evens)
sum(ourlist[["evens"]])

identical(ourlist$evens,ourlist["evens"])

attributes(ourlist)
names(ourlist)

```

# Arrays: creating

* Like matrix, but more than two dimensions
* Only one data type (like vector, matrix)
* Create an array using the array() function

```{r}
vector1 = c(1,3,5)
vector1

vector2 = c(20,21,22,23,24,25)
vector2

#Create two 3x3 matrices
arrayVariable = array(c(vector1,vector2),dim = c(3,3,2))
arrayVariable
```

# Arrays: accessing

```{r}
arrayVariable

#get matrix 1
arrayVariable[,,1]

#get 3rd row of matrix 2
arrayVariable[3,,2]

#get element in row 1, column 3 in matrix 1
arrayVariable[1,3,1]

#get matrix 2
arrayVariable[,,2]

```

# Arrays: named elements

```{r}
#Create the names
rowNames = c("row1","row2","row3")
rowNames

columnNames = c("col1","col2","col3")
columnNames

matrixNames = c("matrix1","matrix2")
matrixNames

arrayVariable = array(c(vector1,vector2), dim = c(3,3,2), dimnames = list(rowNames, columnNames, matrixNames))
arrayVariable

#Access by names:
arrayVariable[,,"matrix1"]
arrayVariable["row2","col1","matrix1"]

```

# Data frames: creating

* Used to store data tables
* List of equal sized vectors
* Create a data.frame using the data.frame() function

```{r}
name = c("patient1", "patient2", "patient3")
name

age = c(46, 49, 50)
age

smoker = c(T, F, T)
smoker

patientRecords = data.frame(name, age, smoker)
patientRecords
```

# Data frames: accessing

```{r}
patientRecords

#Just like matrices
patientRecords[1,] #row 1
patientRecords[,1] #column 1
patientRecords[1,2] #element at row 1, column 2
patientRecords$age

```

# Exercise: Data frames (10 minutes)

>Read the "gene_condition_source_id.txt" file used in Assignment 1 into a data frame variable (see assignment 1 for the command)

>Look at the top few rows using the head() command

>How many rows and columns does the file have? Try the str() command. Also try nrow() and ncol()

>Use names() to find the column names 

### Exercise solution (5 minutes)

```{r}
ourdf <- read.table("../data/gene_condition_source_id.txt",
                    sep="\t",header=T,quote="",comment.char="",
                    as.is=T)
str(ourdf)
head(ourdf)
nrow(ourdf)
ncol(ourdf)
names(ourdf)
colnames(ourdf)
names(ourlist)
rownames(ourdf)[1:5]

sapply(ourdf,is)
ourdf$X.GeneID[1:5]
is(ourdf$AssociatedGenes[1:5])


seq(1,10)[-(1:5)]

```


***
