---
title: "Module 6"
---

The Book of R: chapter 11.1, 11.2 (functions), Glance at ?base::regex

***

# Module 5 review

## Loops

```
for (value in vectorVariable) {
  if (value > 0) {
    print(value)
  }
}
```

## Vectorization

```
vectorVariable[vectorVariable>0]
```   
    
***

## Exercise: Vectorization (10 minutes)

>Create a vector of consecutive numbers from 1 to 100.  
>Note: from:to is the same as seq(from, to)  
>Compute the sum, product and mean  
>Sum just the numbers greater than 50  

### Exercise solution (5 minutes)

```{r}

```

# Functions

## Writing your own functions

```{r}
#syntax
#Note: myfunction can be any name
myfunction = function(arg1, arg2, ... ) {
  statements
  return(object)
}

#Note: objects inside the function are local to the function i.e. not defined outside. This is called variable scope
#The concept of scope is very important in other languages, but in R, you just need to understand it with functions
```

## Function example

```{r}
patientRecords = data.frame(name=c("patient1", "patient2", "patient3"),
                            age=c(46, 49, 50), 
                            smoker=c(T, F, T),
                            stringsAsFactors=F)


printRecordRow = function(patientDatabase) {
	#prints a row in a patient database
	v = paste(patientDatabase[1],
	patientDatabase[2], patientDatabase[3])
	return(v)
}

#apply to one row
printRecordRow(patientRecords[1,])
#apply the printRecordRow function to all the rows of the patientRecords matrix
apply(patientRecords, 1, printRecordRow)


```

## Why create functions?

* Organize your code – useful for larger scripts
* Easy to reuse code because it is more modular
    * Collect your own library of useful functions for easy scripting later (can be stored in a separate file and included in all your scripts)
    * More easily share code with others
* Maintain a central place to keep code in case you need to modify/fix something
* Easily repeat code from different places in a single script

## Built in functions

* R has hundreds of built in functions

```{r}
#e.g. statistical functions
t.test(runif(50), runif(50))
cor.test(runif(50), runif(50),method="pearson")
```

## Exercise: Functions (10 minutes)

>Practice creating a function  
>Create a function called mysum that sums up all the elements of a vector  
>Check that it gives the same results at the sum() function  

### Exercise solution (5 minutes)

```{r}


```


# Regular expressions

* Useful tool to find patterns in strings

* Regular expressions (regex) = very powerful pattern matching in strings

```{r}
string = "aaabbbccc"
#"aaa" matches
#"abb" matches
#"abbbb" does it match?

#You’ve already been using regular expressions e.g.
DNA = "ACGTGA"
sub("A", "T", DNA)
sub("GA", "CT", DNA)
gsub("A", "T", DNA)

```

* There are lots of ways to specify a pattern – it’s a mini language within a language

string = "aaabbbccc"
"a.a" matches "aaa"
. means any one character

"a*b" matches
* means zero or more of the previous character

Example: [+-]?[0-9]+\.[0-9]+
matches a number like -43.221


## Regular expression quick reference

| aaa             | exact pattern                                                                  |
|-----------------|--------------------------------------------------------------------------------|
| .               | any character                                                                  |
| aa\|bb           | aa or bb                                                                       |
| ()              | group a pattern e.g. (y\|n)o matches yo and no                                  |
| []              | character set e.g. [abc] matches a single character, either a, b or c; a-z     |
| [^]             | not character set e.g. [^abc] matches any single character except a, b or c    |
| [:digit:] or \\d | a range of standard character classes are available                            |
| [:alpha:]       |                                                                                |
| *               | 0 or more of the previous character or group                                   |
| +               | 1 or more                                                                      |
| ?               | 0 or 1                                                                         |
| {n,m}           | a specific number of repeats (e.g. 2 to 5).                                    |
|                 | Also {n,} = n to infinity and {n} = exactly n                                  |
| ^               | start of string                                                                |
| $               | end of string or before a \\n                                                   |
| \\\\b             | word boundary e.g. \\bword\\b (\\B matches non word boundaries)                   |
|                 | Also \\< and \\> match start and end of a word                                   |
| \\\\N             | After the regular expression is run:                                           |
|                 | backreference the Nth pattern matched in a ()  (up to 9 patterns)              |
| ignore.case=T   | case insensitive match                                                         |
| \\\\+ \\\\^ \\\\      | ‘escape’ a metacharacter – use a special character like + or ^ in your pattern |

## Regular expression examples

```{r}
a = "sdfsabcdefghba0dsgggfe0haaabbbccc"
#grep returns the vector index of the match
#grepl returns TRUE if there is a match
#regexpr returns a list, including match index and length; gregexpr returns all matches
grep("b.d", a)
grepl("b.d", a)
regexpr("b.d", a)
regexpr("b.d", a)[1]
gregexpr("b.d", a)

# b.d			  #The dot will match any character
# b[ce]d		#the [ce] matches a single 'c' or a single 'e'.
# b[ce]+d	  #the [ce]+ matches any number of c's & e's (at least one)
# b[ce]*d		#the [ce]* also can match the empty string.
# b\\sd		  #the \s matches blank or tab (so-called	'whitespace')
# b\\s+d		#the \s+ matches any number of whitespaces (at least one) 
# a|b+		  #what will this match?
# (a|b)+		#and this?

#Note: to run these yourself, add them to e.g. gregexpr("b.d", a), where "b.d" is the regular expression above and “a” is the string variable to search.

gregexpr("b.d", a)

```

* string = "Kevin Bacon"
* (K|k)evin (B|b)acon
* Start (^) and end ($) of a string
    * ^(Kevin|Bacon|Portman|Roberts)
    * (Kevin|Bacon|Portman|Roberts)$
    * ^(Bacon|Portman|Roberts)
* Find all vowels
    * (A|E|I|O|U|Y|a|e|i|o|u|y)
    * [AEIOUYaeiouy]
* How many matches?

```{r}
length(gregexpr("g", a)[[1]])
```

## Exercise: regular expressions (10 minutes)

>DNA = "ACGATGATACTGATGACGGGC"  
>Match the regular expression "G..A" and use the results to identify the actual match in the original string  
>Use gregexpr  
>attr(x, "match.length") can be used to find the match length in the vector returned from gregexpr  
>?substr  

### Exercise solution (5 minutes)

```{r}

```


***
