---
title: "Module 9"
---

The Book of R: Chapter 7 (plotting)

***

## Module 8 review

* Packages
    * Installing
        * install.packages("kinship2")
        * library(kinship2)
    * Creating your own package
* Bioconductor
    * rich source of bioinformatics packages
    
***

# The reproducible paper

* Journals are moving towards fully reproducible papers, where R code behind each figure is available for anyone to play with

https://repro.elifesciences.org/example.html#

* Plotting in R is very powerful, but takes some getting used to

## The built in R plot() command

```{r}

### Using R's built in plot command
plot(c(1,2,3))
plot(c(1,2,3), c(2,3,4))

#cosine curve
x=seq(from = 0, to = 20, by = 0.05)
x=cos(x)
plot(x)

#plot options

#type: plot type e.g. lines connecting data points
#main, xlab, ylab: title, x and y axis labels
#col: colour options
#pch: point style
#cex: point size
#lty: line style
#lwd: line width
#xlim, ylim: x and y axis range

#adding a title
plot(c(1,2,3), c(2,3,4), main="My plot")

#adding labels
plot(c(1,2,3), c(2,3,4), xlab="points 1", ylab="points 2")

#colours
colors()  #shows all the colours defined by default in R

plot(c(1,2,3), c(2,3,4), col = "red")

#point style
plot(c(1,2,3), c(2,3,4), pch = 19)
plot(c(1,2,3), c(2,3,4), pch = 19, cex = 2)  #point size

# Many point styles available
# http://www.endmemo.com/program/R/pchsymbols.php

#line style
plot(c(1,2,3), c(2,3,4), type = "l")  #line plot
plot(c(1,2,3), c(2,3,4), type = "b")  #line + points
plot(c(1,2,3), c(2,3,4), type = "o")  #plot points on top of the line

#lty can range from 1 to 6
plot(c(1,2,3), c(2,3,4), type = "o", lty = 1)
plot(c(1,2,3), c(2,3,4), type = "o", lty = 2)
plot(c(1,2,3), c(2,3,4), type = "o", lty = 1, lwd = 3)

#axis range
#normally R sets the range automatically, but you can set it manually
plot(c(1,2,3), c(2,3,4), xlim = c(0,5))

```

## Combining parameters to create an advanced plot

```{r}
{ #note: in notebooks, the regular plot window is not used, thus is not active. When legend is called, it will not know where the plot is, so surround code with {} to group all the commands together into one command

# Colour and point style
cols  <- c(rep('red',50), rep('blue',50), rep('green',50))
types <- c(rep(1, 50), rep(2, 50), rep(3, 50))

plot(iris$Petal.Length, iris$Petal.Width,
main='Iris Petal Variability',
xlab='Petal Length (cm)', ylab='Petal Width (cm)',
col=cols, pch=types, cex.axis=1.5, cex.lab=1.5
)

# Adding a legend
legend('topleft', legend=c('I. setosa', 'I. versicolor', 'I. virginica'), col=c('red', 'blue', 'green'), pch=c(1,2,3))
}
```

## Building a plot in stages

```{r}
#R creates a new plot with each call to plot(). Other functions allow you to add plot components to an existing plot

x = 1:10
y = 10:1

{
plot(x,y, type = "n") #type n means empty plot

#abline() adds straight lines
abline(h=c(2.5,3,3.5), lty=2, lwd=2)  #adds horizontal dashed lines

#add points
points(x[y>5],y[y>5], col = "red", pch = 3)
points(x[y<=5],y[y<=5], col = "blue")

#add line
lines(x,y)
}


#you can add a legend using legend() as in the previous example

```


## Exercise: Create a simple plot (10 minutes)

* Plot a quadratic curve including numbers from -50 to 50. Make the negative numbers red and the positive numbers blue, with zero a purple, filled in point. Build the plot in stages.

### Exercise solution (5 minutes)

```{r}
# From https://stackoverflow.com/a/30835971
line2user <- function(line, side) {
  lh <- par('cin')[2] * par('cex') * par('lheight')
  x_off <- diff(grconvertX(c(0, lh), 'inches', 'npc'))
  y_off <- diff(grconvertY(c(0, lh), 'inches', 'npc'))
  switch(side,
         `1` = grconvertY(-line * y_off, 'npc', 'user'),
         `2` = grconvertX(-line * x_off, 'npc', 'user'),
         `3` = grconvertY(1 + line * y_off, 'npc', 'user'),
         `4` = grconvertX(1 + line * x_off, 'npc', 'user'),
         stop("Side must be 1, 2, 3, or 4", call.=FALSE))
}

# pdf("thisisapdf.pdf",width=5,height=5)
# png("thisisnotapdf.png",width=5,height=5,units="in",res=300)
par(mar=c(6,6,3,1),mgp=c(2,1,0))
plot(NA,NA,xlim=range(-50:50),ylim=c(0,50^2))
points(1:50,(1:50)^2,type="l",col="blue")
lines(-50:-1,(-50:-1)^2,col="red")
points(0,0,col="purple",pch=16,cex=2)

# now I'm just being silly

mtext("Farts!",side=3,line=1)
mtext("Farts!",side=3,line=0,las=2)
text(0,2500/2,"Middle!")
text(0,0,"I'm over here",srt=45,xpd=NA)
# par("usr")
text(line2user(0.1,2),line2user(0.1,1),xpd=NA,
     label="I'm over here now!",srt=45,adj=c(1,0.5))
# dev.off()
```


# Exploring other plot types

```{r}
# Read.table yeast_properties.txt (examine the file in a spreadsheet)
yeastdata<-read.table("yeast_properties.txt", header=TRUE)
smoothScatter(yeastdata[,"dnds"], log(yeastdata[,"Degree"]))

#the par() command supports many graphical parameter settings.

#putting two plots side by side
{
par(mfrow=c(1,2))  #2 plots in 1 row
plot(yeastdata[,"dnds"], log(yeastdata[,"Degree"]))
smoothScatter(yeastdata[,"dnds"], log(yeastdata[,"Degree"]))
par(mfrow=c(1,1))  #back to 1 plot in 1 row (more useful in the R console)
}
```

## Example: lattice plot package

```{r}
# http://cran.r-project.org/web/packages/lattice/index.html
# install.packages('lattice')  #installs from CRAN
library(lattice)
```

```{r}
xyplot(yeastdata[,"dnds"] ~ log(yeastdata[,"Degree"]) | Age, data=yeastdata )
```

## ggplot2 package - the grammar of graphics

```{r}
# http://ggplot2.org/
# http://docs.ggplot2.org/current/

# install.packages("ggplot2")
library(ggplot2)
```

```{r}

#quick plot
plot(1:10, 10:1)
qplot(1:10, 10:1)  #notice that ggplot2 looks nicer by default
qplot(1:10, 10:1) + geom_line()  #add a line
qplot(1:10, 10:1) + geom_line(color="red", size=2, linetype=2)  #customize the line style


#and the more powerful ggplot command
ggplot(data=yeastdata, aes(x=yeastdata[,"dnds"], y=yeastdata[,"Degree"])) + 
geom_point() + 
scale_y_log10() + 
xlab("dN/dS") + 
ylab("Degree")

#as you can see, ggplot is more intuitive for building a plot in layers

```

## Circos plots

```{r}
#http://cran.r-project.org/web/packages/RCircos/RCircos.pdf

# install.packages("RCircos")
library(RCircos)
```

```{r fig.height=5, fig.width=5}
#test circos plot
RCircos.Workflow()

data(UCSC.HG19.Human.CytoBandIdeogram)
cyto.info <- UCSC.HG19.Human.CytoBandIdeogram

{
RCircos.Set.Core.Components(cyto.info, chr.exclude=NULL, 10, 0)

RCircos.Set.Plot.Area()

RCircos.Chromosome.Ideogram.Plot()
}

```

## Exercise: Fun with plotting in R (10 minutes)

> Create a 3D scatter plot of dnds, log(degree) and cai from the yeast data table  
> Try using the “highlight.3d” parameter  
> Hint:  
> `install.packages("scatterplot3d")`  
> `library(scatterplot3d)`  


### Exercise solution (5 minutes)

```{r}
yeastdata<-read.table("yeast_properties.txt", header=TRUE)
scatterplot3d::scatterplot3d(yeastdata$dnds,
                             log(yeastdata$Degree),
                             yeastdata$cai,
                             highlight.3d=T)

```



# Assignment 5 hint!
```{r, eval=F}
primer_input <- list()
for (J in 1:length(SEQUENCES)) {
    primer_input[[J]] <- c(paste0("PRIMER_SEQUENCE_ID=",SEQNAME),
                           paste0("SEQUENCE=",toupper(SEQUENCE)),
                           "PRIMER_FILE_FLAG=0",
                           "PRIMER_EXPLAIN_FLAG=1",
                           "=")
}
proper_primer3_input <- unlist(primer_input)
```

